 Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings

### 2020-11-26 paper.Rmd
Comment added in Sec 4
resectioning and spelling

A general difficulty with using 
non-linear DR methods for exploratory data analysis is selecting and tuning 
appropriate parameters. For concretenes here we focus on t-distributed
stochastic neighbour embedding (t-SNE), and we will examine
its underpinning in some detail below [@Maaten2008-sk]. Similar considerations hold for related
methods, for example UMAP [@McInnes2018-co].

@@

Tours provide a supportive visualisation to NLDR graphics, and 
can be easily incorporated into an analysts workflow with our software package, 
\pkg{liminal}.  Our interface allows analysts to quickly compare 
views from embedding methods and see how an embedding method 
preserves or alters the geometry of their data. Using multiple concatenated and
linked views with the tour enhances interaction techniques, and allows 
analysts to perform  cluster orientation tasks via linked highlighting and 
brushing [@McDonald1982-ew; @Becker1987-gd]. 
This approach allows our interface to achieve 
the three principles for interactive high-dimensional data visualisation outlined
by @Buja1996-fk: finding gestalt, posing queries, and making comparisons.

@@

Combining multiple views of a single data set allows the analyst
to make meaniningful comparisons. In \pkg{liminal} the embedding
and tour views are arranged side-by-side for direct cross-checks.

@@

@@

From the video linked inn figure \@ref(fig:gaussian), we learn that t-SNE has 
correctly split out each cluster and laid them out in a star like formation. 
This is agrees with the tour view,
where once we start the animation, the five clusters begin to appear but
generally points are more concentrated in the projection view compared to 
the t-SNE layout (figure \@ref(fig:gaussian)a). 
This can be seen via brushing the t-SNE view (figure \@ref(fig:gaussian)b).

@@

Next we view Gaussian clusters from the 
_Multi Challenge Dataset_, a benchmark simulation data set for clustering tasks
[@Rauber2009-vh]. This dataset has two Gaussian clusters with equal covariance 
embedded in 10-$d$, and a third cluster
with hierarchical structure. This cluster has two 3-$d$ clusters embedded in 
10-$d$, where the second cluster is subdivided into three smaller clusters, 
that are each equidistant from each other and have the same covariance structure.
From the video linked in figure \@ref(fig:hierarchical), we see 
that t-SNE has correctly identified the sub-clusters. However, their relative 
locations to each other is distorted, with the orange and blue groups being
far from each other in the tour view (figure \@ref(fig:hierarchical)a). We see 
in this case that is difficult
to see the sub-clusters in the tour view, however, once we zoom and highlight
they become more apparent (figure \@ref(fig:hierarchical)b). When we brush 
the sub-clusters in the t-SNE, their
relative placement is again exaggerated, with the tour showing that they
are indeed much closer than the impression the t-SNE view gives.

(ref:liminal-hierarchical) Screenshots of the \pkg{liminal} interface applied to sub-clustered data, a video of the tour animation is available at https://player.vimeo.com/video/43963590.

@@

Next we explore some simulated noisy tree structured data (figure
\@ref(fig:fake-trees)), our interest here is how t-SNE visualisations break
topology of the data, and then seeing if we can resolve this by tweaking the
default parameters with reference to the global view of the data set. 
This simulation aims to mimic branching trajectories of cell differentiation:
if there were only mature cells, we would just see the tips of the branches 
which have a hierarchical pattern of clustering.

(ref:fake-trees) Example high-dimensional tree shaped data, $n = 3000$ and $p = 100$. *(a)* The true data lies on 2-$d$ tree consisting of ten branches. This data is available in the \pkg{phateR} package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce]. *(b)* The first two principal components, which form the initial projection for the tour, note that the backbone of the tree is obscured by this view. *(c)* The default t-SNE view breaks the global structure of the tree. *(d)* Altering t-SNE using the first two principal components as the starting coordinates for the embedding, results in clustering the tree at its branching points.

@@

Moreover, we run t-SNE using the default arguments on the complete data (this
keeps the first 50 PCs, sets the perplexity to equal 30 and performs random
initialisation). We then create a linked
tour with t-SNE layout with \pkg{liminal} as shown in figure \@ref(fig:trees-01).

@@

From the linked video, we see that the t-SNE view has been unable to 
recapitulate the topology of the tree - the backbone (blue) branch has been 
split into three fragments (figure \@ref(fig:trees-01)a). We can see 
this immediately via the linked highlighting over both
plots. If we click on the legend for the zero branch, the blue coloured points
on each view are highlighted and the remaining points are made transparent.
From, here it becomes apparent from the tour view that the blue branch forms
the backbone of the tree and is connected to all other branches. From the video 
it is easy to see that cluster sizes formed via t-SNE can be misleading; from the tour
view there is a lot of noise along the branches, while this does not appear to
be the case for the t-SNE result (figure \@ref(fig:trees-01)b).

From the first view, we modify the inputs to the t-SNE view, to try and produce
a better trade-off between local structure and retain the topology of the data.
We keep every parameter the same except that we initialise $Y$ with the first
two PCs (scaled to have standard deviation 1e-4) instead of the default random
initialisation and increase the perplexity from 30 to 100. We then combine
these results with our tour view as displayed in the linked video in the caption
of figure \@ref(fig:trees-02).

@@

The video linked in figure \@ref(fig:trees-02) shows that this selection of 
parameters results in the 
tips of the branches (the three black dots in figure \@ref(fig:fake-trees)a) 
being split into three clusters representing the terminal branches of the tree. 
However, there are perceptual issues following the placement of the three 
groupings on the t-SNE view that become apparent via simple linked brushing. 
If we brush the
tips of the yellow and brown branches (which appear to be close to each other
on the t-SNE view), we immediately see the placement is distorted in the t-SNE
view, and in the tour view these tips are at opposite ends of the tree 
(figure \@ref(fig:trees-02)b).
Although, this is a known issue of the t-SNE algorithm, we can easily identify
it via simple interactivity without knowing the inner workings of the method.

@@

First, we downloaded the raw mouse retinal single cell RNA-seq data from
@Macosko2015-ot using the \pkg{scRNAseq} Bioconductor package [@scRNAseq-d]. We
have followed a standard workflow for pre-processing and normalizing this data 
(described by @Amezquita2020-at): we performed QC using the \pkg{scater} package
by removing cells with high proportion of mitochondrial gene expression and low
numbers of genes detected, we log-transformed and normalised the expression 
values and  finally selected highly variable genes (HVGs) using \pkg{scran} 
[@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
the normalised expression matrix and compute PCA using the first 25 components.
Using the PCs we built a shared nearest neighbours graph (with $k = 10$) and 
used Louvain clustering to generate clusters [@Blondel2008-bx].

To check and verify the clustering we construct a \pkg{liminal} view. 
We tour the first five PCs (approximately 20% of the variance in expression),
alongside the t-SNE view which was computed from all 25 PCs. 
Due to latency of the interface we do a weighted sample of  the rows based on
cluster membership, leaving us with approximately 10 per cent of the original
data size - 4,590 cells. Although this is not ideal, it still allows us to
get a sense of the shape of the clusters as seen from the linked video in figure \@ref(fig:mouse-01).

@@

```{r mouse-01, fig.cap = "(ref:mouse-01)", out.height = "85%"}
display_screenshot("mouse-01", vid_lookup)
```

@@

From the video linked in figure \@ref(fig:mouse-01), we learn that the embedding has mostly captured the
clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
linked brushing (figure \@ref(fig:mouse-01)a). As expected, t-SNE mitigates the crowding problem that is
an issue for tour in this case, where points in clusters 2,4,6, and 11 are
clumped together in tour view, but are blown up in the embedding view (figure \@ref(fig:mouse-01)b). 
The tour appears to form a tetrahedron like shape, with points lying on
surface and along the vertices of the tetrahedron in 5-$d$ PCA space -
a phenomena that has also been observed in @Korem2015-af 
(figure \@ref(fig:mouse-01)c). Brushing on
the tour view, reveals points in cluster 9 that are diffuse in the embedding 
view, points in cluster 9 are relatively far away and spread apart from other 
clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding
(figure \@ref(fig:mouse-01)d).

@@

From the video linked in figure \@ref(fig:mouse-02), we see that the 
expression of the marker genes, appear to separate
the previously clumped clusters 2,4,6, and 11 from the other clusters, 
indicating that these genes are expressed in all four clusters (figure \@ref(fig:mouse-02)a). After zooming, we can see a trajectory forming along
the clusters, while the axis view shows that magnitude of expression in 
the marker genes is similar across these separated clusters which is 
consistent with the results of marker gene analysis (figure \@ref(fig:mouse-02)b).

## Discussion

@@

We have shown in the case studies, that one to one linked brushing can
be used to identify distortions in the embedding, however we would
like extend this to one to many linked brushing, which would allow us to directly interrogate neighbourhood preservation. This form of brushing acts directly on a $k$-nearest neighbours ($k$-nn) graph computed from a
reference dataset: when a user brushes over a region in the embedding,
all the points that match the graphs edges are selected on the 
corresponding tour view. The reference data set for computing
nearest neighbours (for example a distance matrix, or the complete data matrix) 
can be independent of the tour or embedding views. In place of 
highlighting, one  could use opacity or binned colour scales to encode 
distances or ranks instead of the neighbouring points. We have begun 
implementing this brush in \pkg{liminal},  using the  \pkg{FNN} or \pkg{RcppAnnoy}
packages for fast neighbourhood estimation on the server side, however there
are still technicalities that need be resolved [@fnn-pkg; @rcpp-annoy-pkg]. 
Brush composition, such as 'and', 'or', or 'not' brushes, could be used to 
further investigate mismatches between the $k$-nn graphs estimated from both the embedding and tour views.

There are some limitations
in using the \pkg{liminal} interface for larger datasets. First,
t-SNE avoids the crowding problem, points are separated into distinct regions
on the display canvas, while for the tour, points are concentrated in the
centre of the projection and become difficult to see. 
Second, as $n$ increases the animation becomes slower as there is more data
passing from the server to the client. Both of these issues could be mitigated
through alternate tour paths and displays. The recent work
by @Laa2020-wr is a promising area for further investigation, as well as
work from topological statistics [@Rieck2017-kk; @Genovese2017-iq].

>>>>

A general difficulty with using 
non-linear DR methods for exploratory data analysis is selecting and tuning 
appropriate parameters. For concreteness here we focus on t-distributed
stochastic neighbour embedding (t-SNE), and we will examine
its underpinning in some detail below [@Maaten2008-sk]. Similar considerations hold for related
methods, for example UMAP [@McInnes2018-co].

@@

Tours provide a supportive visualisation to NLDR graphics, and 
can be easily incorporated into an analysts workflow with our software package, 
\pkg{liminal}.  Our interface allows analysts to quickly compare 
views from embedding methods and see how an embedding method 
preserves or alters the geometry of their data. Using multiple concatenated and
linked views with the tour enhances interaction techniques, and allows 
analysts to perform  cluster orientation tasks via linked highlighting and 
brushing [@McDonald1982-ew; @Becker1987-gd]. 
This approach allows our interface to achieve 
the three principles for interactive high-dimensional data visualisation outlined
by @Buja1996-fk:  finding gestalt (identifying patterns in visual forms), posing queries, and making comparisons.

@@

Combining multiple views of a single data set allows the analyst
to make meaningful comparisons. In \pkg{liminal} the embedding
and tour views are arranged side-by-side for direct cross-checks.

@@

\textcolor{red}{Ursula: can I use restart to replay the same tour path (i.e. find a
previously seen projection), or will it start a new random selection of bases?}

@@

From the video linked in Figure \@ref(fig:gaussian), we learn that t-SNE has 
correctly split out each cluster and laid them out in a star like formation. 
This is agrees with the tour view,
where once we start the animation, the five clusters begin to appear but
generally points are more concentrated in the projection view compared to 
the t-SNE layout (Figure \@ref(fig:gaussian)a). 
This can be seen via brushing the t-SNE view (Figure \@ref(fig:gaussian)b).

@@

Next we view Gaussian clusters from the 
_Multi Challenge Dataset_, a benchmark simulation data set for clustering tasks
[@Rauber2009-vh]. This dataset has two Gaussian clusters with equal covariance 
embedded in 10-$d$, and a third cluster
with hierarchical structure. This cluster has two 3-$d$ clusters embedded in 
10-$d$, where the second cluster is subdivided into three smaller clusters, 
that are each equidistant from each other and have the same covariance structure.
From the video linked in Figure \@ref(fig:hierarchical), we see 
that t-SNE has correctly identified the sub-clusters. However, their relative 
locations to each other is distorted, with the orange and blue groups being
far from each other in the tour view (Figure \@ref(fig:hierarchical)a). We see 
in this case that is difficult
to see the sub-clusters in the tour view, however, once we zoom and highlight
they become more apparent (Figure \@ref(fig:hierarchical)b). When we brush 
the sub-clusters in the t-SNE, their
relative placement is again exaggerated, with the tour showing that they
are indeed much closer than the impression the t-SNE view gives.

(ref:liminal-hierarchical) Screenshots of the \pkg{liminal} interface applied to sub-clustered data, a video of the tour animation is available at https://player.vimeo.com/video/439635905.

@@

Next we explore some simulated noisy tree structured data (Figure
\@ref(fig:fake-trees)). Our interest here is how t-SNE visualisations break
topology of the data, and then seeing if we can resolve this by tweaking the
default parameters with reference to the global view of the data set. 
This simulation aims to mimic branching trajectories of cell differentiation:
if there were only mature cells, we would just see the tips of the branches 
which have a hierarchical pattern of clustering.

(ref:fake-trees) Example high-dimensional tree shaped data, $n = 3000$ and $p = 100$. *(a)* The true data lies on 2-$d$ tree consisting of ten branches. This data is available in the **phateR** package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce]. *(b)* The first two principal components, which form the initial projection for the tour, note that the backbone of the tree is obscured by this view. *(c)* The default t-SNE view breaks the global structure of the tree. *(d)* Altering t-SNE using the first two principal components as the starting coordinates for the embedding, results in clustering the tree at its branching points.

@@

Moreover, we run t-SNE using the default arguments on the complete data (this
keeps the first 50 PCs, sets the perplexity to equal 30 and performs random
initialisation). We then create a linked
tour with t-SNE layout with \pkg{liminal} as shown in Figure \@ref(fig:trees-01).

@@

From the linked video, we see that the t-SNE view has been unable to 
recapitulate the topology of the tree - the backbone (blue) branch has been 
split into three fragments (Figure \@ref(fig:trees-01)a). We can see 
this immediately via the linked highlighting over both
plots. If we click on the legend for the zero branch, the blue coloured points
on each view are highlighted and the remaining points are made transparent.
From, here it becomes apparent from the tour view that the blue branch forms
the backbone of the tree and is connected to all other branches. From the video 
it is easy to see that cluster sizes formed via t-SNE can be misleading; from the tour
view there is a lot of noise along the branches, while this does not appear to
be the case for the t-SNE result (Figure \@ref(fig:trees-01)b).

From the first view, we modify the inputs to the t-SNE view, to try and produce
a better trade-off between local structure and retain the topology of the data.
We keep every parameter the same except that we initialise $Y$ with the first
two PCs (scaled to have standard deviation 1e-4) instead of the default random
initialisation and increase the perplexity from 30 to 100. We then combine
these results with our tour view as displayed in the linked video in the caption
of Figure \@ref(fig:trees-02).

@@

The video linked in Figure \@ref(fig:trees-02) shows that this selection of 
parameters results in the 
tips of the branches (the three black dots in Figure \@ref(fig:fake-trees)a) 
being split into three clusters representing the terminal branches of the tree. 
However, there are perceptual issues following the placement of the three 
groupings on the t-SNE view that become apparent via simple linked brushing. 
If we brush the
tips of the yellow and brown branches (which appear to be close to each other
on the t-SNE view), we immediately see the placement is distorted in the t-SNE
view, and in the tour view these tips are at opposite ends of the tree 
(Figure \@ref(fig:trees-02)b).
Although, this is a known issue of the t-SNE algorithm, we can easily identify
it via simple interactivity without knowing the inner workings of the method.

@@

First, we downloaded the raw mouse retinal single cell RNA-seq data from
@Macosko2015-ot using the **scRNAseq** Bioconductor package [@scRNAseq-d]. We
have followed a standard workflow for pre-processing and normalizing this data 
(described by @Amezquita2020-at): we performed QC using the **scater** package
by removing cells with high proportion of mitochondrial gene expression and low
numbers of genes detected, we log-transformed and normalised the expression 
values and  finally selected highly variable genes (HVGs) using **scran** 
[@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
the normalised expression matrix and compute PCA using the first 25 components.
Using the PCs we built a shared nearest neighbours graph (with $k = 10$) and 
used Louvain clustering to generate clusters [@Blondel2008-bx].

To check and verify the clustering we construct a \pkg{liminal} view. 
We tour the first five PCs (approximately 20% of the variance in expression),
alongside the t-SNE view which was computed from all 25 PCs. We have selected only the first five PCs because there is a large drop in the percentage of variance explained after the fifth component, with each component after contributing less than one percent of variance.
Consequently, increasing the number of PCs to tour would increase the dimensionality and volume of the subspace we are touring but without adding any additional signal to the view.

Due to latency of the \pkg{liminal} interface we do a weighted sample of  the rows based on
cluster membership, leaving us with approximately 10 per cent of the original
data size - 4,590 cells. Although this is not ideal, it still allows us to
get a sense of the shape of the clusters as seen from the linked video in Figure \@ref(fig:mouse-01). If one was interested in performing more in-depth cluster analysis we recommend an iterative approach of removing large clusters and then re-running the \pkg{liminal} view as a way finding more granular cluster structure.

@@

```{r mouse-01, fig.cap = "(ref:mouse-01)", out.height = "90%"}
display_screenshot("mouse-01", vid_lookup)
```

@@

From the video linked in Figure \@ref(fig:mouse-01), we learn that the embedding has mostly captured the
clusters relative location to each other to their location in high dimensional space, with a notable exception of points in cluster 3 and 10 as shown with
linked brushing (Figure \@ref(fig:mouse-01)a). As expected, t-SNE mitigates the crowding problem that is
an issue for tour in this case, where points in clusters 2,4,6, and 11 are
clumped together in tour view, but are blown up in the embedding view (Figure \@ref(fig:mouse-01)b). 
The tour appears to form a tetrahedron like shape, with points lying on
surface and along the vertices of the tetrahedron in 5-$d$ PCA space -
a phenomena that has also been observed in @Korem2015-af 
(Figure \@ref(fig:mouse-01)c). Brushing on
the tour view, reveals points in cluster 9 that are diffuse in the embedding 
view, points in cluster 9 are relatively far away and spread apart from other 
clusters in the tour view, but has points placed in cluster 3 and 9 in the embedding
(Figure \@ref(fig:mouse-01)d).

@@

From the video linked in Figure \@ref(fig:mouse-02), we see that the 
expression of the marker genes, appear to separate
the previously clumped clusters 2,4,6, and 11 from the other clusters, 
indicating that these genes are expressed in all four clusters (Figure \@ref(fig:mouse-02)a). After zooming, we can see a trajectory forming along
the clusters, while the axis view shows that magnitude of expression in 
the marker genes is similar across these separated clusters which is 
consistent with the results of marker gene analysis (Figure \@ref(fig:mouse-02)b).

# Discussion

@@

We have shown in the case studies, that one to one linked brushing can
be used to identify distortions in the embedding, however we would
like extend this to one to many linked brushing, which would allow us to directly interrogate neighbourhood preservation. This form of brushing acts directly on a $k$-nearest neighbours ($k$-nn) graph computed from a
reference dataset: when a user brushes over a region in the embedding,
all the points that match the graphs edges are selected on the 
corresponding tour view. The reference data set for computing
nearest neighbours (for example a distance matrix, or the complete data matrix) 
can be independent of the tour or embedding views. In place of 
highlighting, one  could use opacity or binned colour scales to encode 
distances or ranks instead of the neighbouring points. We have begun 
implementing this brush in \pkg{liminal},  using the  **FNN** or **RcppAnnoy**
packages for fast neighbourhood estimation on the server side, however there
are still technicalities that need be resolved [@fnn-pkg; @rcpp-annoy-pkg]. 
Brush composition, such as 'and', 'or', or 'not' brushes, could be used to 
further investigate mismatches between the $k$-nn graphs estimated from both the embedding and tour views.

There are some limitations
in using the \pkg{liminal} interface for larger datasets. First,
t-SNE avoids the crowding problem, points are separated into distinct regions
on the display canvas. For the tour, points are concentrated in the
centre of the projection and become difficult to see. We have recently proposed a simple non-linear transformation for the tour called a sage tour that aims to fix this problem [@Laa2020-vk]. 
Second, as $n$ increases both the embedding view and tour view become harder to read due to over-plotting, while the interactivity and animation become slower as there is more data passing from the server to the client.
For the tasks we have looked at in this paper, where shape and density are important to the analyst, we think that better displays and sub-sampling strategies are more useful than being able to look at every single point on the canvas. We showed in our single cell clustering case study that doing a weighted sample based on cluster membership still allowed us to get a sense of relative cluster orientation, however there are alternative sampling approaches that could be applied, like selecting points close to the cluster centres. Alternative displays via statistical transformations could also mitigate the need to show all of the data. Recent work
by @Laa2020-wr is a promising area for further investigation, as well as
work from topological statistics [@Rieck2017-kk; @Genovese2017-iq].
